\LARGE \bf Preparation of Papers for IEEE Sponsored Conferences \& Symposia*

### 2017-02-08 pwarel.tex
.

>>>>

\section{Relational Modeling}

We now describe how relational models can be computed for a given
black box system by enriching the abstract reachability graph obtained
by scatter-and-simulate. We first formalize the notion of an enriched
graph, and then show how they can be interpreted as a PWA transition
system using $k$-relational modeling.

\subsection{Abstract Enriched Graph}

The existential abstraction relation in \chapref{abs} was defined as
follows:
\[
%\exists \x\in\C. \exists \x'\in\C'. \x'=\simulate(\x, \tau) \iff \C\rel{\tau}\C'
    C \areach{t} C' \iff \exists \x \in C.\ \exists \x' \in C'.\ \x \areach{t} \x'
\]
The abstract relation $\areach{t}$ can be enriched by incorporating
the affine relation between $\x$ and $\x'$. For an arbitrary dynamical
system, such a relation can be rarely represented using an exact affine map.
This is due to the presence of non-linear and hybrid behaviors. But,
an affine map can always be estimated with an error.

If the system dynamics are completely specified in the form of a white
box model, we can use first order approximations to find the affine
expressions for the relations. Using a tool like \flowstar, we can
obtain sound over-approximate affine maps of the form $A\x + [\vb^l,
\vb^h]$, where $[\vb^l,\vb^h]$ denotes the interval of vectors, such
that, every element $b_i$ of the vector $\vb$ is contained in the
respective scalar interval $b_i\in[b^l_i,b^h_i]$.

For the case of black box systems, such a sound approximation is not
possible. Instead, we rely on a statistical method like simple linear
regression to estimate the affine map $\amap: A\x + \vb$. We then
estimate the error $\delta$ and generalize $\amap$ to an interval
affine map as before $\amap: A\x + \vb + \delta$.  Finally, we get an
abstraction with the below relation
\[
    C \areach{f} C' \iff \exists \x \in C.\ \exists \x' \in C'.\ \x' \in A\x + \vb \pm \delta.
    %C \areach{A\x+\vb+\delta} C' \iff \exists \x \in C.\ \exists \x' \in C'.\
\]

However, in the presence of complex non-linear behavior, using a
single affine map can lead to a poor approximation. To increase the
precision, one can use more than one affine map. Let us denote this set as
$R_{(C,C')}$ (ref.~\secref{ovw}).
\[
    R_{(C,C')}(\x,\x'): \setof{f_i \;|\; \exists \x \in C.\ \exists \x' \in C'.\ \x' \in f_i(\x)}
\]

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}

    \draw [line] (-1.2,-1.2) rectangle (1.2,1.2);
    \draw[arw] (-3,0) -- (-1.2,0);
    \draw[arw] (1.2,0) -- (3,0);

\node at (-3.4,0) {$G$};
\node at (3.4,0) {$G^R$};
\node[align=center] at (0, 0) {\scriptsize Enrich using \\ \scriptsize Regression \\ \scriptsize(OLS)};

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{Using OLS, $G^R$ is computed by determining the appropriate
$R_{(C,C')}$ for each edge of $G$.}
\label{fig:enrichment}
\vspace*{-.3cm}
\end{figure}

As shown in \figref{enrichment}, we use OLS to compute an enrichment
of the abstraction graph $G$. For every edge $(C,C') \in edges(G)$,
the set of relations $R_{(C,C')}(\x,\x')$ is computed. The semantics
of the enriched edge are clearly non-deterministic. Form a state $\x
\in C$, any $f_i\in R_{(C,C')}$ can be taken as long as $\x' \in C$'
is satisfied. This interpretation results in an infinite state
transition system. We now detail the construction of the set
$R_{(C,C')}$ using $k$-relational modeling.

% We now compute a PWA model for a given black box system by estimating
% the dynamics using OLS. We can estimate the affine relations between
% every cell, but as discussed in ~\chapref{abs}, it is a futile
% approach.  Instead, we use the same heuristic as before:
% scatter-and-simulate, to select the cell relations for estimation. We
% build the reachability graph as before, but in addition, annotate each
% edge with the values of estimated $A$, $\vb$, and $\delta$ for the
% respective relation. The reachability graph thus computed, is a
% transition system. Using off-the-shelf bounded model checkers, we
% reason about the system's safety properties.

\subsection{$k$ - Relational Modeling}

\begin{figure}[!htbp]
\vspace*{-.2cm}
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}

\begin{scope}[scale=0.8]
\draw [line] (-1.0,0) circle (0.5);
\draw [line] (-4.0,0) circle (0.5);
\draw [line] (2.0,1) circle (0.5);
\draw [line] (2.0,-1) circle (0.5);
\draw[arw] (-0.5,0) -- (1.5,1.0);
\draw[arw] (-0.5,0) -- (1.5,-1.0);
\draw[arw] (-3.5,0) -- (-1.5,0.0);
\node at (-4.0,0) {$C$};
\node at (-1.0,0) {$C'$};
\node at (2.0,1.0) {$C''_1$};
\node at (2.0,-1.0) {$C''_2$};
\draw [line] (2.0,-1) circle (0.5);
\draw[arw] (-5.0,-2.0) -- (3.0,-2.0);
\node at (-4.0,-2.5) {$k=0$};
\node at (-1.0,-2.5) {$k=1$};
\node at (+2.0,-2.5) {$k=2$};
\draw [fill=black] (-4.0,-2.0) circle (0.05);
\draw [fill=black] (-1.0,-2.0) circle (0.05);
\draw [fill=black] (2.0,-2.0) circle (0.05);

\end{scope}
\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{Nodes/Cells of $G$ shown along with increasing values of $k$.}
\label{fig:k-rel}
\vspace*{-.3cm}
\end{figure}

Given a $G$, when computing the set of relations for an edge $R$, we
consider only the set of abstract states reachable at a specific time
step.  Intuitively, one might consider the states reachable in one
time step $\Delta$.  However, such a process can be generalized by
looking at the states reachable at ${0\Delta, 1\Delta, \ldots,
k\Delta}$ time steps.

Using \figref{k-rel}, we illustrate this notion. To compute
$R_{(C,C')}$, we
split the data set $D(C,C')$ consisting of trajectory segments (of time lengths
$k\Delta$) as follows.  We observe the local behavior of the system by
noting the evolution of the system $\System$ from a state $\x(t)
\in C$ for a time length dependant on $k$. For
\begin{itemize}
    \item{$k=0$}: we observe all trajectory segments of length $\Delta$.
    \item{$k=1$}: we observe trajectory segments of length $\Delta$, which satisfy $\x(t+\Delta) \in C'$.
    \item{$k=2$}: we observe trajectory segments of length $2\Delta$,
        and split them in two sets (a) and (b) on the basis of the
        cells reached at time $(t+\Delta)$ and $(t+2\Delta)$.
        \begin{enumerate}[(a)]
            \item $\x(t+\Delta) \in C' \land \x(t+2\Delta) \in C''_1$
            \item $\x(t+\Delta) \in C' \land \x(t+2\Delta) \in C''_2$
        \end{enumerate}
    \item{$k=n$}: we observe trajectory segments of length $n\Delta$,
        and split them in to multiple sets on the basis of the cells
        reached at time $(t+\Delta),\; (t+2\Delta),\;
        \ldots,\;(t+n\Delta)$.
\end{itemize}

$k$-relational modeling can be understood as a heuristic, which uses
the underlying abstraction to differentiate behaviors of the system
which `diverge' (or are revealed to be `distinct' at a future time).
Such a heuristic can be useful in increasing the precision of the
learnt affine maps.  We now formalize this notion for $k=0$, $1$, and
$n$ and illustrate using examples.

\mypara{$0$-Relational Model}

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}
\begin{scope}[scale=0.5]

    \draw [line] (-1.0,-1.0) rectangle (1.0,1.0);
    %\draw [line] (3.0,1.0) rectangle (5.0,3.0);
    %\draw [line] (3.0,-3.0) rectangle (5.0,-1.0);
    \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,2.2);
    \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,1.7);
    \draw [line] (-0.0,-0.5) .. controls +(2.0,2.0) and +(-2.0,-1.0) ..  (3.5,-1.2);

\node at (0.0, 0.0) {$C$};
\node at (-0.75,0.7) {\scriptsize{$\x_0$}};
\node at (3.7,2.2) {\scriptsize{$\x_0'$}};
\node at (-0.7,0.0) {\scriptsize{$\x_1$}};
\node at (3.7,1.7) {\scriptsize{$\x_1'$}};
\node at (-0.2,-0.5) {\scriptsize{$\x_2$}};
\node at (3.7,-1.2) {\scriptsize{$\x_2'$}};
\node at (2,1.5) {$\pi_1$};
\node at (2,0.5) {$\pi_2$};
\node at (1.7,-1) {$\pi_3$};
\draw[arw] (-1.5,-2.0) -- (4.5,-2.0);
\node at (0,-2.5) {$t=0$};
\node at (4,-2.5) {$t=\Delta$};
\draw [fill=black] (0.0,-2.0) circle (0.05);
\draw [fill=black] (4.0,-2.0) circle (0.05);

\end{scope}
\begin{scope}[xshift=0,yshift=3.0cm,scale=0.5]
    \node at (2,0) {\footnotesize $\scr{T} \equiv g_1(\x):\vx \in C
    \implies A_1\vx + \vb_1 + \delta_1$};
\end{scope}

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{All the trajectory segments will be used to construct the
model; $\ds = \setof{\pi_1, \pi_2, \pi_3}$.}
\label{fig:k0}
\vspace*{-.3cm}
\end{figure}

When $k=0$, the $0$-relational model is a PWA transition system which only
defines the evolution of states $\x$, and does not specify the
reachable cell. The guard predicates of the relations are defined over
cells: $g_i(\x): \x \in C_i$ while $g'_i(\x'): True$.

We use regression to estimate the dynamics for the outgoing trajectory
segments form a cell $C$. Hence, the data set $\ds$ for the regression
includes all trajectory segments beginning from the same cell
\[
    \ds = \setof{\pi_t | start(\pi_t) \in C}.
\]
This includes trajectory segments ending in different cells, as shown
in ~\figref{k0}. For $N$ cells, this results in a $\rho$ with $N$ transition relations.
\begin{equation*}
    \scr{T} = \left\{
        \begin{array}{ll}
            g_1(\x):\vx \in C_1 \implies A_1\vx + \vb_1 + \delta_1\\
            \ldots \\
            g_n(\x):\vx \in C_n \implies A_n\vx + \vb_n + \delta_n\\
        \end{array}
    \right.
\end{equation*}

Note that this can be quite imprecise when the cells are big,
containing regions of state-space with complex dynamics. This is true
for both non-linear systems and hybrid dynamical system, where a cell
can contain two or more modes with differing continuous dynamics.

\mypara{$1$-Relational Model}

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}
\begin{scope}[scale=0.5]

    \draw [line] (-1.0,-1.0) rectangle (1.0,1.0);
    \draw [line] (3.0,1.0) rectangle (5.0,3.0);
    \draw [line] (3.0,-3.0) rectangle (5.0,-1.0);
    \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,2.2);
    \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,1.7);
    \draw [line] (-0.0,-0.5) .. controls +(2.0,2.0) and +(-2.0,-1.0) ..  (3.5,-1.2);

\node at (0.0, 0.0) {$C$};
\node at (4.0, 2.0) {$C'_1$};
\node at (4.0, -2.0) {$C'_2$};
\node at (-0.75,0.7) {\scriptsize{$\x_0$}};
\node at (-0.7,0.0) {\scriptsize{$\x_1$}};
\node at (-0.2,-0.5) {\scriptsize{$\x_2$}};
\node at (3.3,2.5) {\scriptsize{$\x_0'$}};
\node at (3.7,1.5) {\scriptsize{$\x_1'$}};
\node at (3.7,-1.5) {\scriptsize{$\x_2'$}};
\node at (2,1.5) {$\pi_1$};
\node at (2,0.5) {$\pi_2$};
\node at (1.7,-1) {$\pi_3$};
\draw[arw] (-1.5,-3.5) -- (5.5,-3.5);
\node at (0,-4.0) {$t=0$};
\node at (4,-4.0) {$t=\Delta$};
\draw [fill=black] (0.0,-3.5) circle (0.05);
\draw [fill=black] (4.0,-3.5) circle (0.05);

\end{scope}
\begin{scope}[xshift=0,yshift=4.0cm,scale=0.5]
    \node at (2,0) {\begin{minipage}{\linewidth}
    \footnotesize \begin{equation*}
        \scr{T} = \left\{
        \begin{array}{ll}
            g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_1\vx + \vb_1 + \delta_1\\
            g_2(\x):\vx \in C \land g'_2(\x):\vx \in C'_2 \implies A_2\vx + \vb_2 + \delta_2
        \end{array}
    \right.
\end{equation*}
\end{minipage}};
\end{scope}

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{The data gets split into two sets $\ds_1 = \setof{\pi_1,
    \pi_2}$ and $\ds_2 = \setof{\pi_3}$ and two relations:
    $R_{(C,C'_1)}$ and $R_{(C,C'_2)}$, each with one affine map, are constructed.}
%Cardinality($R_{(C,C')}$) = $2$.}
\label{fig:k1}
\vspace*{-.3cm}
\end{figure}

% \begin{scope}[xshift=0,yshift=4.0cm,scale=1]
%     \node at (2,0) {\footnotesize $\begin{equation*}
%     \scr{T} = \left\{
%         \begin{array}{ll}
%             g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_1\vx + \vb_1 + \delta_1\\
%             g_2(\x):\vx \in C \land g'_2(\x):\vx \in C'_2 \implies A_2\vx + \vb_2 + \delta_2
%         \end{array}
%     \right.
% \end{equation*}$};
% \end{scope}

To improve the preciseness of learnt dynamics, we include the
reachability relation in the
regression. For every relation $C\areach{t}C'$, the data set $\ds$ is
comprised only of trajectory segments $\pi_t$ which start and end in
the same set of cells.
\[
    \ds = \setof{\pi_t | start(\pi_t) \in C \land end(\pi_t) \in C'}.
\]
For $N$ edges in $G$, $1$-relationalization results in a $\rho$ with
$N$ transition relations.

\figref{k1} illustrates the $k=1$ refinement of the case shown in
\figref{k0}. The data set $\ds$ is split into two data sets $\ds_1
= \setof{\pi_1, \pi_2}$ and $\ds_2 = \setof{\pi_3}$, using which, two
relations are constructed. The resulting $1$-relational model
is at least as precise as the corresponding $0$-relational model.

\mypara{$k$-Relational Model}

\begin{figure}[!htbp]
\begin{center}
\tikzstyle{line} = [thick]
\tikzstyle{arw} = [->, thick,>=stealth,shorten <=2pt, shorten >=2pt]
\begin{tikzpicture}
\begin{scope}[scale=0.5]

    \draw [line] (-1.0,-1.0) rectangle (1.0,1.0);
    \draw [line] (3.0,1.0) rectangle (5.0,3.0);
    %\draw [line] (3.0,-3.0) rectangle (5.0,-1.0);
    \draw [line] (7.0,1.0) rectangle (9.0,3.0);
    \draw [line] (7.0,-3.0) rectangle (9.0,-1.0);
    \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (3.5,2.2) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (8.5,1.8) ;
    \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (3.5,1.7) .. controls +(2.0,2.0) and +(-2.0,-2.0) .. (8.5,-2.2) ;

%     \draw [line] (-0.7,0.5) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,2.2);
%     \draw [line] (-0.5,0.0) .. controls +(2.0,2.0) and +(-2.0,-2.0) ..  (3.5,1.7);

    %\draw [line] (-0.0,-0.5) .. controls +(2.0,2.0) and +(-2.0,-1.0) ..  (3.5,-1.2);

\node at (0.0, 0.0) {$C$};
\node at (4.0, 2.0) {$C'_1$};
\node at (8.0, 2.0) {$C''_1$};
\node at (8.0, -2.0) {$C''_2$};
%\node at (4.0, -2.0) {$C'_2$};
\node at (-0.75,0.7) {\scriptsize{$\x_0$}};
\node at (-0.7,0.0) {\scriptsize{$\x_1$}};
%\node at (-0.2,-0.5) {\scriptsize{$\x_2$}};
\node at (3.3,2.5) {\scriptsize{$\x_0'$}};
\node at (3.7,1.5) {\scriptsize{$\x_1'$}};
%\node at (3.7,-1.5) {\scriptsize{$\x_2'$}};
\node at (8.7,1.6) {\scriptsize{$\x_0''$}};
\node at (8.5,-2.7) {\scriptsize{$\x_1''$}};
\node at (2,1.5) {$\pi_1$};
\node at (2,0.5) {$\pi_2$};
\draw[arw] (-1.5,-3.5) -- (9.5,-3.5);
\node at (0,-4.0) {$t=0$};
\node at (4,-4.0) {$t=\Delta$};
\node at (8,-4.0) {$t=2\Delta$};
\draw [fill=black] (0.0,-3.5) circle (0.05);
\draw [fill=black] (4.0,-3.5) circle (0.05);
\draw [fill=black] (8.0,-3.5) circle (0.05);
\end{scope}

\begin{scope}[xshift=0,yshift=4.0cm,scale=0.5]
    \node at (4,0) {\begin{minipage}{\linewidth}
        \footnotesize \begin{equation*}
    \scr{T} = \left\{
        \begin{array}{ll}
            g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_1\vx + \vb_1 + \delta_1\\
            g_1(\x):\vx \in C \land g'_1(\x):\vx \in C'_1 \implies A_2\vx + \vb_2 + \delta_2
        \end{array}
    \right.
\end{equation*}
\end{minipage}};
\end{scope}

\end{tikzpicture}
\end{center}
\vspace*{-.3cm}
\caption{The $k=2$ refinement further splits the data set $\ds_1$
    into two sets $\ds_{11} = \setof{\pi_1}$ and $\ds_{12} =
    \setof{\pi_2}$ and $R_{(C,C'_1)}$ now has two affine maps,
    non-deterministically defining the system behavior.}
\label{fig:k2}
\vspace*{-.3cm}
\end{figure}

Finally, we generalize the relational models to $k$-relational PWA
models. A $k$-relational model is constructed by using $k$ length
\textit{connected} segmented trajectories. A segmented trajectory $\SegTraj$ is
\textit{connected} iff its cost $\cost(\SegTraj) = 0$.

Two $\SegTraj:\tupleof{\pi_{t_1}, \ldots, \pi_{t_k}}$ and
$\SegTraj':\tupleof{\pi'_{t_1}, \ldots, \pi'_{t_k}}$ are
\textit{similar} if their trajectory segments have the same sequence
of cell traversals $\tupleof{C_1, C_2, \ldots, C_k}$, \ie
\[
    \forall i\in\setof{1 \ldots k}.
    start(\pi_{t_i}) \in C_i
    \land start(\pi'_{t_i}) \in C_i
    \land end(\pi_{t_i}) \in C_{i+1}
    \land end(\pi'_{t_i}) \in C_{i+1}
\]

For every cell $C$ in the reachability graph, we construct a data set
$\ds$ by collecting $\pi_{t_i}$, such that they are the first segment of $k$
length \textit{connected} and \textit{similar} segmented trajectories.
%\[
%    \ds = \setof{\pi_{\tau_i} |
%        \pi_{\tau_i}
%        start(\pi_{\tau_{i}}) \in C_i
%        \land
%        end(\pi_{\tau_{i}}) \in C_i}
%\]

\figref{k2} illustrates the $k=2$ refinement on \figref{k1}.
